\bf Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-20 appendix.qmd
More described the approach

@fig-pbmc-mse-umap shows layout from the original published article along with that suggested by scDEED to be more accurate. The log MSE vs binwidth plot illustrates that our approach would suggest that scDEED is correct here, that layout b is more accurately reflecting the cluster structure in the data.

@@

error_umap_plot_pbmc <- ggplot(error_pbmc,
                               aes(x = a1,
                                   y = MSE,
                                   colour = method)) +
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) +
  # geom_vline(xintercept = 15, linetype="solid",
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the 2 UMAP layouts with different hyper-parameter settings (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

>>>>

The layout a (@fig-pbmc-mse-umap) is generated from the hyper-parameters suggested by @chen2023, and the layout b (@fig-pbmc-mse-umap) is with suggested hyper-parameters by scDEED to be more accurate. The layout a and b contain $46$ and $15$ dubious cells respectively.

The log MSE vs binwidth ($a_1$) plot (@fig-pbmc-mse-umap) illustrates that our approach would suggest that scDEED is correct here, that layout b is more accurately reflecting the cluster structure in the PBMC data.

@@

error_umap_plot_pbmc <- ggplot(error_pbmc,
                               aes(x = a1,
                                   y = MSE,
                                   colour = method)) +
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the two UMAP layouts with different hyper-parameter settings (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $80$, min_dist: $0.5$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP layout (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

### 2025-04-20 appendix.tex
More described the approach

Figure~\ref{fig-pbmc-mse-umap} shows layout from the original published
article along with that suggested by scDEED to be more accurate. The log
MSE vs binwidth plot illustrates that our approach would suggest that
scDEED is correct here, that layout b is more accurately reflecting the
cluster structure in the data.

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyper-parameter settings (n\_neighbors: \(30\),
min\_dist: \(0.3\) (red); n\_neighbors: \(80\), min\_dist: \(0.5\)
(blue)) on the PBMC data is the better representation using MSE for
varying binwidth (\(a_1\)). Colour used for the lines and points in the
top plot and in the scatterplots represents UMAP layout (a, b). Of the
two, layout b is optimal across all binwidths making it the best
choice.}

>>>>

The layout a (Figure~\ref{fig-pbmc-mse-umap}) is generated from the
hyper-parameters suggested by \citet{chen2023}, and the layout b
(Figure~\ref{fig-pbmc-mse-umap}) is with suggested hyper-parameters by
scDEED to be more accurate. The layout a and b contain \(46\) and \(15\)
dubious cells respectively.

The log MSE vs binwidth (\(a_1\)) plot (Figure~\ref{fig-pbmc-mse-umap})
illustrates that our approach would suggest that scDEED is correct here,
that layout b is more accurately reflecting the cluster structure in the
PBMC data.

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the two UMAP
layouts with different hyper-parameter settings (n\_neighbors: \(30\),
min\_dist: \(0.3\) (red); n\_neighbors: \(80\), min\_dist: \(0.5\)
(blue)) on the PBMC data is the better representation using MSE for
varying binwidth (\(a_1\)). Colour used for the lines and points in the
top plot and in the scatterplots represents UMAP layout (a, b). Of the
two, layout b is optimal across all binwidths making it the best
choice.}
